Efficiently Handling Guards in the Low Level Design of RPython's Tracing JIT

### 2012-08-08 talk/vmil2012/paper.tex
remove a redundant todo
some improvements

The core idea of storing resume data as compactly as possible
is to share parts of the data structure between subsequent guards.
This is often useful because the density of guards in traces is so high,
that quite often not much changes between them.
Since resume data is a linked list of symbolic frames
often only the information in the top frame changes from one guard to the next.
The other frames can often be just reused.
The reason for this is that during tracing only the variables
of the currently executing frames can change.
Therefore if two guards are generated from code in the same function
the resume data of the rest of the stack can be reused.

@@

Guards interact with optimizations in various ways.
Most importantly optimizations try to remove as many operations
and therefore guards as possible.
This is done with many classical compiler optimizations.
In particular guards can be removed by subexpression elimination.
If the same guard is encountered a second time in the trace,
the second one can be removed.
This also works if a later guard is weaker implied by a earlier guard.

@@

Adding this additional information is done as follows.
So far, every variable in the symbolic frames
contains a constant or an SSA variable.
After allocation removal the variables in the symbolic frames can also contain
``virtual'' objects.
These are objects that were not allocated so far,
because the optimizer removed their allocation.
The virtual objects in the symbolic frames describe exactly
how the heap objects that have to be allocated on guard failure look like.
To this end, the content of every field of the virtual object is described
in the same way that the local variables of symbolic frames are described.
The fields of the virtual objects can therefore be SSA variables, constants
or other virtual objects.
They are encoded using the same compact two-byte representation
as local variables.

@@

Similarly, stores into the heap are delayed as long as possible.
This can make it necessary to perform these delayed stores
when leaving the trace via a guard.
Therefore the resume data needs to contain a description
of the delayed stores to be able to perform them when the guard fails.
So far no special compression is done with this information.

@@

\todo{compare to naive variant of resume data}
\todo{Measure the of guards and how many of these ever fail}

>>>>

The core idea of storing resume data as compactly as possible
is to share parts of the data structure between subsequent guards.
This is often useful because the density of guards in traces is so high,
that quite often not much changes between them.
Since resume data is a linked list of symbolic frames
often only the information in the top frame changes from one guard to the next.
The other frames can often be just reused.
The reason for this is that during tracing only the variables
of the currently executing frame can change.
Therefore if two guards are generated from code in the same function
the resume data of the rest of the frame stack can be reused.

@@

Guards interact with optimizations in various ways.
Most importantly optimizations try to remove as many operations
and therefore guards as possible.
This is done with many classical compiler optimizations.
In particular guards can be removed by subexpression elimination.
If the same guard is encountered a second time in the trace,
the second one can be removed.
This also works if a later guard is weaker
and therefore implied by a earlier guard.

@@

Adding this additional information is done as follows.
So far, every variable in the symbolic frames
contains a constant or an SSA variable.
After allocation removal the variables in the symbolic frames can also contain
``virtual'' objects.
These are objects that were not allocated so far,
because the optimizer removed their allocation.
The virtual objects in the symbolic frames describe exactly
how the heap objects look like which have to be allocated on guard failure.
To this end, the content of every field of the virtual object is described
in the same way that the local variables of symbolic frames are described.
The fields of the virtual objects can therefore be SSA variables, constants
or other virtual objects.
They are encoded using the same compact two-byte representation
as local variables.

@@

A related optimization is the handling of heap stores by the optimizer.
The optimizer tries to delay stores into the heap as long as possible.
This is done because often heap stores become unnecessary
due to another store to the same memory location later in the trace.
This can make it necessary to perform these delayed stores
when leaving the trace via a guard.
Therefore the resume data needs to contain a description
of the delayed stores to be able to perform them when the guard fails.
So far no special compression is done with this information,
compared to the other source of information delayed heap stores are quite rare.

@@

\todo{compare to naive variant of resume data}
